---
title: "Population Data Management Lab"
author: "Alicia Rich"
output:
  html_document:
    theme:
      bootswatch: litera
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: "show"
    fig_caption: true
    df_print: paged
params:
  course: "zoobio"
  semester: "spring25"
  studbook_current: "Studbooks/Demo/Studbook_Living.csv"
  studbook_historic: "Studbooks/Demo/Studbook_Historic.csv"
  
---

```{r, include = FALSE}
library(knitr)
library(conflicted)
library(downloadthis)
library(fontawesome)
library(glue)
library(ggplot2)
library(ggtext)
library(heatmaply)
library(here)
library(htmltools)
library(kableExtra)
library(kinship2)
library(paletteer)
library(pedtools)
library(pedsuite)
library(plotly)
library(reactable)
library(reactablefmtr)
library(rmarkdown)
library(scales)
library(shiny)
library(showtext)
library(thematic)
library(tidyverse)
library(tippy)
library(usethis)

conflicts_prefer(here::here)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::left_join)
conflicts_prefer(dplyr::inner_join)
conflicts_prefer(dplyr::full_join)
conflicts_prefer(dplyr::semi_join)
conflicts_prefer(dplyr::rename)
conflicts_prefer(ggplot2::margin)
conflicts_prefer(ggplot2::theme_classic)
conflicts_prefer(ggplot2::ggplot)
conflicts_prefer(ggplot2::theme_minimal)
conflicts_prefer(ggplot2::aes)
conflicts_prefer(dplyr::lag)
conflicts_prefer(purrr::set_names)
conflicts_prefer(purrr::flatten)
conflicts_prefer(purrr::discard_at)
conflicts_prefer(base::which.max)
conflicts_prefer(lubridate::month)
conflicts_prefer(lubridate::year)
conflicts_prefer(lubridate::day)
conflicts_prefer(base::as.data.frame)
conflicts_prefer(htmltools::p)
conflicts_prefer(purrr::discard)
```


```{r, include = FALSE}
here::i_am("PopulationData.Rmd")


theme_set(theme_classic())
thematic_rmd()
thematic_on(accent = "#8785B2FF", fg = "black")

opts_chunk$set(message = FALSE,
               warning = FALSE,
               echo    = TRUE,
               include = TRUE,
               eval    = TRUE,
               comment = "")

palette <- c("#D53288FF", "#DC8045FF", "#21B14BFF", "#008AC2FF", "#3F459BFF")

colors <- list(
  f = "#D53288FF",
  m = "#3F459BFF",
  u = "#21B14BFF",
  sire = "#3F459B33",
  dam  = "#D5328833",
  seq  = "rcartocolor::Sunset",
  div  = "rcartocolor::Temps",
  rand = "Redmonder::qPBI"
)

deceased.col <- function(color) {
  gsub("FF", "33", color)
}


font_add_google("Noto Sans Symbols", family = "NotoSym")
showtext_auto()



```

# Format Studbook Data

## Inspect Spreadsheets in Excel

First, we need to read in our studbook data and clean it up to create a *tidy* format of one master table.

1.  Open the file `Studbook_Living.csv` and make sure it has the following columns with the exact names listed.

- It is fine if you have extra columns, and all the columns may be in any order, but your spreadsheet must include at least these columns with the exact name syntax:
  - `ID`
    - *This is the studbook ID (not the LocalID, if your spreadsheet has one)*
  - `Sex`
  - `Sire`
  - `Dam`
  - `Event_Type`
    - *This column will have values like `Birth/Hatch` or `Transfer`*
  - `Date`
    - *This will be the date for the event on each row. If you have another column for something like birth date, just make sure this general date column is the only one with this name.*
  - `Location`
    - *The same comments from Date apply here. This should be the location corresponding to the event of each row.*
- Rename any columns as necessary, re-save as a `.csv`, and proceed.

2.  Repeat the same for the file `Studbook_Historic.csv` using the same list of columns.

## Read and Format Data in R

```{r}
studbook.living <- read.csv(here("Studbooks/Demo/Studbook_Living.csv"), header = T) %>%
  filter(str_detect(Location, "\\w+")) %>%
  select(ID, Sex, Sire, Dam, Event_Type, Date, Location)  %>%
  mutate(Date  = str_squish(str_extract(Date, ".{1,3}[/-].{1,3}[/-]\\d{1,4}"))) %>%
  mutate(Date  = str_remove_all(Date, "\\s")) %>%
  mutate(Date  = str_replace(Date, "-(?=0\\d$)", "-20"),
         Date  = str_replace(Date, "-(?=[5-9]\\d$)", "-19")) %>%
  mutate(Date  = if_else(str_detect(Date, "^\\d{1,}"), dmy(Date), mdy(Date))) %>%
  mutate(across(where(is.character), ~ na_if(., "")),
         across(where(is.character), ~ str_trim(.)),
         across(where(is.character), ~ str_remove_all(., "\n")),
         Location = str_remove_all(Location, "[^\\w]"),
         across(where(is.character), ~ str_remove_all(., "\\s")),
         ID   = as.integer(str_remove_all(ID  , "[^\\d+]"))) %>%
  fill(ID, Sex, Sire, Dam) %>%
  group_by(ID) %>%
  mutate(Event_order = row_number(), Location = str_to_upper(Location)) %>% 
  fill(Date, .direction = "up") %>%
  ungroup() %>%
  select(Event_order, ID, Sex, Sire, Dam, Event_Type, Date, Location)

studbook.historic <- read.csv(here("Studbooks/Demo/Studbook_Historic.csv"), header = T) %>%
  filter(str_detect(Location, "\\w+")) %>%
  select(ID, Sex, Sire, Dam, Event_Type, Date, Location)  %>%
  mutate(Date  = str_squish(str_extract(Date, ".{1,3}[/-].{1,3}[/-]\\d{1,4}"))) %>%
  mutate(Date  = str_remove_all(Date, "\\s")) %>%
  mutate(Date  = str_replace(Date, "-(?=0\\d$)", "-20"),
         Date  = str_replace(Date, "-(?=[5-9]\\d$)", "-19")) %>%
  mutate(Date  = if_else(str_detect(Date, "^\\d{1,}"), dmy(Date), mdy(Date))) %>%
  mutate(across(where(is.character), ~ na_if(., "")),
         across(where(is.character), ~ str_trim(.)),
         across(where(is.character), ~ str_remove_all(., "\n")),
         Location = str_remove_all(Location, "[^\\w]"),
         across(where(is.character), ~ str_remove_all(., "\\s")),
         ID   = as.integer(str_remove_all(ID  , "[^\\d+]"))) %>%
  fill(ID, Sex, Sire, Dam) %>%
  group_by(ID) %>%
  mutate(Event_order = row_number(), Location = str_to_upper(Location)) %>% 
  fill(Date, .direction = "updown") %>%
  ungroup() %>%
  select(Event_order, ID, Sex, Sire, Dam, Event_Type, Date, Location) %>%
  filter(!(ID %in% studbook.living$ID))
```


```{r}
births <-  studbook.living %>%
  bind_rows(studbook.historic) %>% distinct() %>%
  filter(str_detect(Event_Type, "Birth|Hatch|Capture")) %>%
  arrange(ID, Event_order) %>%
  group_by(ID) %>% 
  arrange(Date, .by_group = TRUE) %>%
  fill(Date, .direction = "downup") %>% 
  ungroup() %>%
  arrange(ID) %>%
  fill(Date, .direction = "downup") %>%
  group_by(ID) %>% slice_min(Event_order) %>%
  ungroup() %>% arrange(Location, ID, Dam) %>%
  mutate(Sire = case_when(
    Sire == "UND" & Dam != "UND" & Location == lag(Location) & lag(Sire) != "UND" ~ lag(Sire),
    Sire == "UND" & Dam != "UND" & Location != lag(Location) & Location == lead(Location) & lead(Sire) != "UND" ~ lead(Sire),
    .default = Sire
  )) %>%
  arrange(Dam) %>%
  mutate(Sire = case_when(
    Sire == "UND" & Dam != "UND" & Location == lag(Location) & lag(Sire) != "UND" ~ lag(Sire),
    Sire == "UND" & Dam != "UND" & Location != lag(Location) & Location == lead(Location) & lead(Sire) != "UND" ~ lead(Sire),
    .default = Sire
  )) %>%
  arrange(Location, ID) %>%
  mutate(Sire = as.integer(str_extract(Sire, "\\d+")),
         Dam  = as.integer(str_extract(Dam, "\\d+"))) %>%
  select(ID, Sex, Sire, Dam, Birth_Date = Date, Birth_Location = Location)
```

```{r}
transfers.deaths  <-  studbook.living %>%
  bind_rows(studbook.historic) %>% distinct() %>%
  filter(str_detect(Event_Type, "Transfer|LTF|Death")) %>%
  arrange(ID, Event_order) %>%
  group_by(ID) %>% fill(Date, .direction = "downup") %>%
  slice_max(Event_order) %>% ungroup() %>%
  mutate(Status = if_else(Event_Type == "Transfer", "Alive", "Deceased")) %>%
  select(ID, Last_Date = Date, Last_Location = Location, Status)

locations <- c(studbook.historic$Location, 
                     studbook.living$Location,
                     births$Birth_Location, 
                     transfers.deaths$Last_Location) %>%
  unique() %>% 
  enframe(name = NULL, value = "Location") %>%
  arrange(Location) %>%
  mutate(loc_num = row_number())
```

```{r}
studbook <- births %>%
  left_join(transfers.deaths) %>%
  distinct() %>%
  mutate(Status        = replace_na(Status, "Alive"),
         Last_Date     = if_else(is.na(Last_Date)    , Birth_Date    , Last_Date),
         Last_Location = if_else(is.na(Last_Location), Birth_Location, Last_Location)) %>%
  mutate(Age = if_else(Status == "Alive", 
                       year(today())   - year(Birth_Date), 
                       year(Last_Date) - year(Birth_Date))) %>%
  arrange(ID) %>%
  mutate(sex_num = case_when(Sex == "Male"   ~ 1,
                             Sex == "Female" ~ 2,
                             .default = 0),
         color = case_when(Sex == "Female"          ~ colors$f,
                           Sex == "Male"            ~ colors$m,
                           Sex == "Undetermined"    ~ colors$u
                           ),
         yr_birth = year(Birth_Date),
         yr_last  = year(Last_Date)
         ) %>%
  left_join(locations, by = join_by(Last_Location == Location)) %>%
  rename(last_loc_num = loc_num) %>%
  left_join(locations, by = join_by(Birth_Location == Location)) %>%
  rename(birth_loc_num = loc_num)


```


## View Formatted File

Let's take a glimpse of our formatted studbook now.

```{r}
studbook.tbl <- studbook %>%
  mutate(Sire = replace_na(Sire, 0),
         Dam  = replace_na(Dam, 0)) %>%
  select(
    Status, 
    ID,      
    Sex,
    color,
    Birth_Date,
    yr_birth,
    Birth_Location,
    birth_loc_num,
    Sire  ,     
    Dam   ,
    Age,
    Last_Date,
    yr_last,
    Last_Location,
    last_loc_num
  ) %>% 
  studbook.react(., studbook.cols(.), groupBy = "Status")
studbook.tbl
```


# Population Viability Analysis

## Demographic Trends

Let's look at the basic population size over time to begin quantifying growth rate and assessing overall population viability.

```{r}
max_month <- floor_date(today(), "month")
max_year  <- floor_date(today(), "year")

alive_months <- pmap(
  list(Birth_Date = studbook$Birth_Date, Status = studbook$Status, Last_Date = studbook$Last_Date),
  function(Birth_Date, Last_Date, Status) {
    
    if (is.na(Birth_Date)) return(NULL)
    
    start_month <- floor_date(Birth_Date, "month")

        if (Status == "Alive") {
          
      end_month <- max_month
      
    } else if (is.na(Last_Date) & Status == "Deceased") {
      return(NULL)
    } else {
      
      end_month <- floor_date(Last_Date, "month")
    }
    # Return the sequence of years from birth to death (inclusive)
    as.list(seq(start_month, end_month, by = "months"))
    
  }
)

monthly_counts <- list_flatten(alive_months) %>%
  list_c() %>%
  tibble(Date = .)  %>%
  group_by(Date) %>%
  summarize(Population = n()) %>% ungroup()

plot.population <- plot_ly(monthly_counts, type = 'scatter', mode = 'lines')%>%
  add_trace(x = ~Date, y = ~Population)%>%
  plotly::layout(showlegend = F, title='Population Growth over Time',
         xaxis = list(rangeslider = list(visible = T)))
plot.population <- plot.population %>%
  plotly::layout(
         xaxis = list(zerolinecolor = '#ffff',
                      zerolinewidth = 2,
                      gridcolor = 'ffff'),
         yaxis = list(zerolinecolor = '#ffff',
                      zerolinewidth = 2,
                      gridcolor = 'ffff'),
         plot_bgcolor='#e5ecf6', width = 900)


plot.population
```

Now we will zoom out to annual population counts to follow the standard method for computing a normalized rate of growth over the past 5 years compared to all years.

```{r}
alive_years <- pmap(
  list(Birth_Date = studbook$Birth_Date, Status = studbook$Status, Last_Date = studbook$Last_Date),
  function(Birth_Date, Last_Date, Status) {
    
    if (is.na(Birth_Date)) return(NULL)
    
    start_year <- floor_date(Birth_Date, "year")

        if (Status == "Alive") {
          
      end_year <- max_year
      
    } else if (is.na(Last_Date) & Status == "Deceased") {
      return(NULL)
    } else {
      
      end_year <- floor_date(Last_Date, "year")
    }
    # Return the sequence of years from birth to death (inclusive)
    as.list(seq(start_year, end_year, by = "years"))
    
  }
)

annual_counts <- list_flatten(alive_years) %>%
  list_c() %>%
  tibble(Year = .)  %>%
  group_by(Year) %>%
  summarize(Count = n()) %>% ungroup()


growth_rates <- annual_counts$Count[-1] / annual_counts$Count[-length(annual_counts$Count)]

last_5_growth <- tail(growth_rates, 5)

lambda  <- prod(last_5_growth)^(1/length(last_5_growth))

lambda2 <- exp(mean(log(last_5_growth)))
```


## Construct Pedigree

Now we will use a package called `pedtools` to create our pedigree objects. If the chunk below runs without any errors, then you should see `pedig` appear as an object listed in your environment.

```{r}
pedig <- ped(
  id     = studbook$ID,
  fid    = studbook$Sire,
  mid    = studbook$Dam,
  sex    = studbook$sex_num
)
```

The object `pedig` most likely contains a list of multiple pedigree objects, as any founders or lineages not represented in the current living population will form discrete, disconnected trees. There also will probably be several "singletons" which are founders without any connections to the living population.\n
\n
The code below will prepare some formatting variables to assign color-coding. Then it will filter out any singletons in the pedigree list and produce a series of images for you to inspect the visual version(s) and choose the most informative image.

```{r}
living   <- filter(studbook, Status == "Alive")    %>% pull(ID) %>% as.character()
deceased <- filter(studbook, Status == "Deceased") %>% pull(ID) %>% as.character()

founders    <- founders(pedig)
nonfounders <- nonfounders(pedig)

female   <- living %>% intersect(females(pedig))
male     <- living %>% intersect(males(pedig))
undet    <- living %>% setdiff(males(pedig)) %>% setdiff(females(pedig))

female.d   <- deceased %>% intersect(females(pedig))
male.d     <- deceased %>% intersect(males(pedig))
undet.d    <- deceased %>% setdiff(males(pedig)) %>% setdiff(females(pedig))

pedig.main <- discard(pedig, \(x) pedsize(x) <= 1)
```

```{r}
annot <- as.list(rep("\u24BB", times = length(founders)))
names(annot) <- founders

fills <- list(
  c(female),
  c(male),
  c(undet)
)

names(fills) <- keep_at(colors, c("f", "m", "u"))

ped.fills <- list(
  "#D5328870" = c(female.d),
  "#3F459B70" = c(male.d),
  "#21B14B70" = c(undet.d)
)

fills <- list_assign(fills, !!!ped.fills)

plot_pedigrees <- function(pedigree, name) {
plot(pedigree,
     title        = paste0(name),
     cex          = 0.4,
     deceased     = deceased,
     labs         = NULL,
     fill         = fills, 
     lwd          = 0.3,
     col          = "black",
     pty          = "m",
     textAnnot    = list(top = list(c(annot), cex = 0.5, family = "NotoSym")))
}

plots <- imap(pedig.main, \(x, idx) plot_pedigrees(x, idx))

```


Inspect these and add the title text for only those pedigree plots including living individuals to the chunk below. Then run the chunk below that to create your main working pedigree object.

```{r}
includes_living <- c("_comp1")

ped.living  <- keep_at(pedig.main, includes_living)

plot_pedigrees(ped.living, "Pedigree of Living Population")

```

## Generate Detailed Pedigree Statistics

Next, we will generate some of the descriptive statistics that we can use to assess the long-term viability of our populations and make some breeding recommendations.

### Founders Represented

We will need information about founders still represented in the living population, so run the code below.

```{r}
founder.descendants <- map(as.list(founders(pedig)), \(x) as.list(descendants(pedig, x, inclusive = TRUE))) %>%
  compact()
names(founder.descendants) <- map(founder.descendants, \(x) x[[1]])

founderReps <- keep(founder.descendants, function(sublist) {
  flat <- unlist(sublist)
  any(flat %in% living)
}) %>% list_flatten(name_spec = "") %>% unique() %>%
  intersect(founders)

n_founder_reps <- length(founderReps)

founderRepsIDs <- founderReps %>% list_c()

dp <- descentPaths(ped.living[[1]])

founderContribution <- map_dbl(founders, function(f) {
  paths_list <- dp[[f]]
  valid_paths <- keep(paths_list, ~ tail(.x, 1) %in% living)
  sum(map_dbl(valid_paths, ~ 0.5^(length(.x) - 1)))
})
names(founderContribution) <- founders

p <- founderContribution / sum(founderContribution)

p.tbl <- enframe(p, name = "ID", value = "Rel_Contribution") %>% mutate(ID = as.integer(ID))
```


```{r}
founder.summary <- enframe(founderContribution, 
                           name = "ID", 
                           value = "Contribution") %>%
  mutate(ID = as.integer(ID)) %>%
  left_join(p.tbl) %>%
  left_join(studbook) %>%
  select(
         ID, 
         Birth_Location, 
         Sex, 
         Birth_Date, 
         Age_Death = Age, 
         Death     = Last_Date, 
         Last_Location,
         Rel_Contribution,
         color         ,
         yr_birth      ,
         yr_last       ,
         birth_loc_num ,
         last_loc_num  
         )  %>%
  arrange(birth_loc_num, desc(Rel_Contribution))

founder.vis <- founder.summary %>%
  studbook.react(., cols = founder.cols(.))

founder.vis
```

### Summary of Living Population

Let's take a quick look at some basic stats for each individual in our living population before we proceed with the overall summary stats.

```{r}
kinship.ped <- ribd::kinship(ped.living[[1]])
living.ped  <- intersect(living, rownames(kinship.ped))
living.kinship <- kinship.ped[living.ped, living.ped]

gen_numbers <- generations(ped.living[[1]], what = "indiv")

living_gen  <- gen_numbers[names(gen_numbers) %in% living]

ancestors <- as.list(rownames(living.kinship)) %>%
  set_names(map(., \(x) x)) %>%
  map(., \(x) as.list(ancestors(ped.living[[1]], x))) %>%
  enframe(name = "ID", value = "Ancestors")

children <- as.list(rownames(living.kinship)) %>%
  set_names(map(., \(x) x)) %>%
  map(., \(x) as.list(children(ped.living[[1]], x))) %>%
  enframe(name = "ID", value = "Children")

descendants <- as.list(rownames(living.kinship)) %>%
  set_names(map(., \(x) x)) %>%
  map(., \(x) as.list(descendants(ped.living[[1]], x))) %>%
  enframe(name = "ID", value = "Descendants")

siblings <- as.list(rownames(living.kinship)) %>%
  set_names(map(., \(x) x)) %>%
  map(., \(x) as.list(siblings(ped.living[[1]], x))) %>%
  enframe(name = "ID", value = "Siblings")

generations <- enframe(gen_numbers, name = "ID", value = "Generations")

living.data <- ancestors %>% 
  left_join(descendants) %>% 
  left_join(children) %>% 
  left_join(siblings) %>%
  left_join(generations) %>%
  rowwise() %>%
  mutate(N_Children    = length(Children),
         N_Descendants = length(Descendants),
         N_Siblings    = length(Siblings),
         N_Ancestors   = length(Ancestors),
         N_Generations = length(Generations)) %>%
  select(ID, starts_with("N_")) %>%
  mutate(ID = as.integer(ID)) %>%
  left_join(studbook) %>%
  select(Current_Location = Last_Location,
         Sex,
         ID, 
         Age,
         Sire,
         Dam,
         Birth_Date,
         Birth_Location,
         starts_with("N_")) %>%
  arrange(Current_Location, Sex, Age)
```


```{r}
living.summary <- living.data %>%
  mutate(
  color = case_when(
    Sex == "Male"           ~ "#332288FF",
    Sex == "Female"         ~ "#AA4499FF",
    Sex == "Undetermined"   ~ "#117733FF"
  )) %>%
  reactable(
    theme   = flatly(),
    defaultPageSize     = 25,
    height              = 700,
    sortable            = TRUE,
    resizable           = TRUE,
    showPageSizeOptions = TRUE,
    columnGroups        = list(
      colGroup(name = "Cummulative Counts", columns = c(
        "N_Ancestors", "N_Children", "N_Siblings"
      ))
    ),
    columns             = list(
      Current_Location = colDef(maxWidth = 100,  name = "Current Location",
                      style = JS("function(rowInfo, column, state) {
                        const firstSorted = state.sorted[0]
                        if (!firstSorted || firstSorted.id === 'Current_Location') {
                          const prevRow = state.pageRows[rowInfo.viewIndex - 1]
                          if (prevRow && rowInfo.values['Current_Location'] === prevRow['Current_Location']) {
                            return { visibility: 'hidden' }
                          }
                        }
                      }")
                      ),
       ID               = colDef(maxWidth = 80, 
                              cell = color_tiles(
                                data       = .,
                                color_ref  = "color",
                                box_shadow = TRUE,
                                bold_text  = TRUE
                                )
                              ),
      Sex   = colDef(show = FALSE),
      N_Generations = colDef(show = FALSE),
      N_Ancestors   = colDef(name  = "Ancestors", 
                             maxWidth = 250,
                             align    = "center", 
                             cell     = bubble_grid(
                               data   = .,
                               opacity = 0.8,
                               colors = paletteer_d("wesanderson::Zissou1Continuous")
                               )
                             ),
      N_Descendants = colDef(name  = "Descendants", 
                             maxWidth = 250,
                             align = "center", 
                             cell  = bubble_grid(
                               data   = .,
                               opacity = 0.8,
                               colors = paletteer_d("wesanderson::Zissou1Continuous")
                               )),
      N_Children    = colDef(name = "Offspring", 
                             maxWidth = 250,
                             align = "center", 
                             cell  = bubble_grid(
                               data    = .,
                               opacity = 0.8,
                               colors  = paletteer_d("wesanderson::Zissou1Continuous")
                               )),
      N_Siblings    = colDef(name = "Siblings", 
                             maxWidth = 250,
                             align = "center", 
                             cell  = bubble_grid(
                               data   = .,
                               opacity = 0.8,
                               colors = paletteer_d("wesanderson::Zissou1Continuous")
                               )),
      
       Age              = colDef(maxWidth = 50),  
      Sire             = colDef(name = "Father",
                                 maxWidth = 80,
                              cell = color_tiles(
                                data       = .,
                                colors      = "#332288FF",
                                opacity     = 0.6,
                                box_shadow = TRUE
                                )
                              ),
      Dam             = colDef(name = "Mother",
                                 maxWidth = 80,
                              cell = color_tiles(
                                data       = .,
                                colors      = "#AA4499FF",
                                opacity     = 0.6,
                                box_shadow = TRUE
                                )
                              ),
      Birth_Date        = colDef(name = "Date of Birth", 
                                 cell = color_tiles(
                                   data       = .,
                                   colors     =  paletteer_d("RColorBrewer::BuPu"),
                                   box_shadow = TRUE,
                                   number_fmt = label_date_short()
                                 ), maxWidth = 200), 
       Birth_Location   = colDef(name = "Location of Birth"),
      color = colDef(show = FALSE)
    )
    
  )

living.summary
```

### Overall Relatedness and Kinship

Now we will use some of the demographic stats we already computed to begin modeling our expected patterns in relatedness and inbreeding across the living population. First, we need to create a matrix representing every living pair's kinship coefficient. We will return to these pairwise relationships soon, but for now we will use it to compute our population-level statistics.

```{r}
mean_gen    <- mean(living_gen, na.rm = TRUE)

F_vec <- 2 * diag(kinship.ped) - 1
F_mean <- mean(F_vec)

delta_F <- 1 - (1 - F_mean)^(1/mean_gen)

Ne <- 1 / (2 * delta_F)

N <- length(living)
Ne_over_N <- Ne / N

FGE <- 1 / sum(p^2)

GD <- 1 - sum(p^2)

MK <- mean(living.kinship[upper.tri(living.kinship)])

F_vec <- 2 * diag(living.kinship) - 1

F_mean <- mean(F_vec)
```

#### Visualize Summary Table

Now you can view a summary of the statistics we just calculated and compare them to the summaries provided in the Pygmy Loris BTP. Use this table and the figures compiled so far to write your own version of a summary (1-2 paragraphs max) like the example.

```{r}
pva.summary <- tibble(N, 
                      n_founder_reps, 
                      FGE, 
                      delta_F, 
                      Ne, 
                      Ne_over_N, 
                      mean_gen, 
                      lambda2, 
                      F_mean, 
                      MK, 
                      GD) %>%
  mutate(across(where(is.numeric), ~round(., digits = 3))) %>%
  reactable(
    theme = espn(),
    columns = list(
      N                = colDef(header = tippy("N"          , tooltip = "Demographic Population size: Number of currently living individuals in the population"                          )), 
      Ne               = colDef(header = tippy("Ne"         , tooltip = "Effective population size (estimated from the rate of inbreeding per generation)"                            ),
                                format = colFormat(digits = 0)), 
      Ne_over_N        = colDef(header = tippy("Ne/N"       , tooltip = "Effective population size relative to population size"),
                                format = colFormat(digits = 1)), 
      n_founder_reps   = colDef(header = tippy("Founders"   , tooltip = "Number of founders represented in the current population"                                 )), 
      FGE              = colDef(header = tippy("FGE"        , tooltip = "Founder genome equivalents: expressed in units of the number of wild-caught, unrelated individuals (“founders”) that would produce the same gene diversity as the current population"                           ),
                                format = colFormat(digits = 1)), 
      F_mean           = colDef(header = tippy("F"          , tooltip = "Mean inbreeding level"                                      )), 
      mean_gen         = colDef(header = tippy("Generations", tooltip = "Mean generation for living individuals"               ),
                                format = colFormat(digits = 1)), 
      delta_F          = colDef(header = tippy("ΔF"         , tooltip = "ΔF is the per–generation increase in inbreeding"                    )), 
      GD               = colDef(header = tippy("GD %"       , tooltip = "Gene diversity: current percentage of founding gene diversity retained"                                       ),
                                format = colFormat(percent = TRUE)), 
      MK               = colDef(header = tippy("MK"         , tooltip = "Population mean kinship"                              )), 
      lambda2          = colDef(header = tippy("lambda"     , tooltip = "Growth rate: Finite rate of population growth calculated as the geometric mean of changes in total population size over the past 5 years"                                          ))
    )
  )

pva.summary
```

## Pairwise Relationships

### All Living Individuals

Now let's look more closely at pairwise patterns in relatedness across our living population. The code below will generate a table with one row for every potential living pair in our population. Keep in mind that these are not necessarily potential breeding pairs, as we are mixing and matching male-male, male-female, and female-female pairs to get an over all impression of the living population's kinship patterns.

```{r}
coeff_living <- coeffTable(ped.living[[1]], coeff = c("f", "phi", "deg", "kappa")) %>%
  filter(id1 %in% living & id2 %in% living) %>%
  select(id1, id2, degree = deg, inbred = f2, phi, kappa0 = k0, kappa1 = k1, kappa2 = k2)

coeff_living
```

We can also generate a demo table with expected values for typical patterns of kinship to help us interpret our results.

```{r}
example <- basicRelationships %>%
  select(label, relationship, phi, starts_with("kappa"))

example
```

It is much easier to visualize pairwise relationships as matrices or multidimensional plots. Let's create some graphical visualizations to interpret these stats. The plot below will give you a basic visualization of the overall patterns in relatedness between all living individuals in the current population, starting with the kinship coefficients that we computed before.

```{r}
kinship.ped    <- ribd::kinship(ped.living[[1]])
living.ped     <- intersect(living, rownames(kinship.ped))
living.kinship <- kinship.ped[living.ped, living.ped]

kin.plot <- heatmaply(
        living.kinship,
        dendrogram     = "none",
        main           = "Kinship Across Living Population",
        scale            = "none",
        colors           = paletteer_d("wesanderson::Zissou1Continuous"),
        margins          = c(60, 100, 40, 20),
        grid_color       = "white",
        grid_width       = 0.00001,
        label_format_fun = function(value) round(value, digits = 3),
        titleX           = TRUE,
        hide_colorbar    = FALSE,
        key.title        = "Kinship Coefficients",
        branches_lwd     = 0.1,
        fontsize_row     = 10, 
        fontsize_col     = 10,
        labCol           = colnames(living.kinship),
        labRow           = rownames(living.kinship),
        heatmap_layers   = theme(axis.line = element_blank())
        )

kin.plot
```


Now we can create similar visuals for the other pairwise coefficients that we calculated.

#### Inbreeding Coefficients

```{r}
inbred.w <- coeff_living %>%
  pivot_wider(id_cols     = "id1",
              names_from  = "id2",
              values_from = "inbred")

inbred.mat <- as.matrix(inbred.w[,-1])
rownames(inbred.mat) <- inbred.w$id1

inbred.mat[lower.tri(inbred.mat)] <- t(inbred.mat)[lower.tri(inbred.mat)]

inbred.plot <- heatmaply(
        inbred.mat,
        dendrogram     = "none",
        main           = "Inbreeding Across Living Population",
        scale            = "none",
        colors           = paletteer_d("wesanderson::Zissou1Continuous"),
        margins          = c(60, 100, 40, 20),
        grid_color       = "white",
        grid_width       = 0.00001,
        label_format_fun = function(value) round(value, digits = 3),
        titleX           = TRUE,
        hide_colorbar    = FALSE,
        key.title        = "Inbreeding Coefficients",
        branches_lwd     = 0.1,
        fontsize_row     = 10, 
        fontsize_col     = 10,
        labCol           = colnames(inbred.mat),
        labRow           = rownames(inbred.mat),
        heatmap_layers   = theme(axis.line = element_blank())
        )

inbred.plot
```

### Matched Pairs

Finally, we want to look closer at potential breeding pairs to do some match-making, so let's generate some more pedigree statistics per individual and then visualize a plot of breeding pairs.

#### Kinship Coefficients

```{r}
living.m       <- filter(studbook, Status == "Alive" & Sex == "Male") %>% 
  mutate(ID = as.character(ID)) %>% 
  pull(ID)

living.f       <- filter(studbook, Status == "Alive" & Sex == "Female") %>% 
  mutate(ID = as.character(ID)) %>% 
  pull(ID)

annotate <- select(living.data, c(ID, starts_with("N_"))) %>%
  mutate(ID = as.character(ID)) %>%
  left_join(enframe(F_vec, name = "ID", value = "F_vec")) %>%
  mutate(ID = as.integer(ID)) %>%
  left_join(studbook) %>%
  mutate(ID = as.character(ID),
    hoverText = str_glue(
    "{Last_Location}", "<br>", 
    "{Age}", " yrs (Born ", "{Birth_Date}", ")<br>",
    "Mother: ", "{Dam}", ", Father: ", "{Sire}", "<br>",
    "{N_Siblings}"," Siblings, ", "{N_Children}", " Offspring, ", "{N_Descendants}", " Descendants"
    )
  ) %>% arrange(ID)


living_males   <- intersect(living.m, rownames(living.kinship))
living_females <- intersect(living.f, rownames(living.kinship))

kinship_btp    <- living.kinship[living_males, living_females]

text_males   <- intersect(living.m, rownames(kinship_btp))
text_females <- intersect(living.f, colnames(kinship_btp))

annotate.m   <- filter(annotate, Status == "Alive" & Sex == "Male") %>% 
  select(ID, hoverText) %>%
  filter(ID %in% text_males) %>% 
  arrange(match(ID, text_males))

annotate.f   <- filter(annotate, Status == "Alive" & Sex == "Female") %>% 
  select(ID, hoverText) %>%
  filter(ID %in% text_females) %>% 
  arrange(match(ID, text_females))

hover_matrix <- outer(
  annotate.m$hoverText, 
  annotate.f$hoverText, 
  FUN = function(m_text, f_text) {
    I(paste0("<br>Male:<br>", m_text, "<br><br>",
           "Female:<br>", f_text))
  }
)

rownames(hover_matrix) <- annotate.m$ID
colnames(hover_matrix) <- annotate.f$ID

btp.plot <- heatmaply(
        kinship_btp,
        dendrogram       = "none", 
        xlab             = "Females", 
        ylab             = "Males",
        main             = "Kinship Values for Potential Mate Pairs",
        scale            = "none",
        colors           = paletteer_d("wesanderson::Zissou1Continuous"),
        margins          = c(60, 100, 40, 20),
        grid_color       = "white",
        grid_width       = 0.00001,
        label_format_fun = function(value) round(value, digits = 3),
        titleX           = TRUE,
        hide_colorbar    = FALSE,
        key.title        = "Kinship Vals",
        branches_lwd     = 0.1,
        fontsize_row     = 10, 
        fontsize_col     = 10,
        custom_hovertext = I(hover_matrix),
        heatmap_layers   = theme(axis.line = element_blank())
        )

btp.plot
```

#### Inbreeding Coefficients

```{r}
living_males   <- intersect(living.m, rownames(inbred.mat))
living_females <- intersect(living.f, rownames(inbred.mat))

inbred_btp    <- inbred.mat[living_males, living_females]

text_males   <- intersect(living.m, rownames(inbred_btp))
text_females <- intersect(living.f, colnames(inbred_btp))

annotate.m   <- filter(annotate, Status == "Alive" & Sex == "Male") %>% 
  select(ID, hoverText) %>%
  filter(ID %in% text_males) %>% 
  arrange(match(ID, text_males))

annotate.f   <- filter(annotate, Status == "Alive" & Sex == "Female") %>% 
  select(ID, hoverText) %>%
  filter(ID %in% text_females) %>% 
  arrange(match(ID, text_females))

hover_matrix <- outer(
  annotate.m$hoverText, 
  annotate.f$hoverText, 
  FUN = function(m_text, f_text) {
    I(paste0("<br>Male:<br>", m_text, "<br><br>",
           "Female:<br>", f_text))
  }
)

rownames(hover_matrix) <- annotate.m$ID
colnames(hover_matrix) <- annotate.f$ID


inbred.plot <- heatmaply(
        inbred_btp,
        dendrogram       = "none", 
        xlab             = "Females", 
        ylab             = "Males",
        main             = "Inbreeding Coefficients for Potential Mate Pairs",
        scale            = "none",
        colors           = paletteer_d("wesanderson::Zissou1Continuous"),
        margins          = c(60, 100, 40, 20),
        grid_color       = "white",
        grid_width       = 0.00001,
        label_format_fun = function(value) round(value, digits = 3),
        titleX           = TRUE,
        hide_colorbar    = FALSE,
        key.title        = "Inbreeding Vals",
        branches_lwd     = 0.1,
        fontsize_row     = 10, 
        fontsize_col     = 10,
        custom_hovertext = I(hover_matrix),
        heatmap_layers   = theme(axis.line = element_blank())
        )

inbred.plot
```

# Inspect Matches

Once you choose some potential pairs, you should plug their ID's into the code chunks below. This will give you a few more graphics to inspect their relationship. You can easily swap out different IDs and re-run the code chunks to see those revised results until you find the pairs you are happy with.

```{r}
male   <- c("2652") # Replace the number here with the ID number of your male of interest.
female <- c("2677") # Replace the number here with the ID number of your female of interest.

pair <- c(male, female)

cat("Your pair's relationship:\n")
verbalise(ped.living, ids = pair)
```

## Kinship Coefficients

```{r}
celltext <- matrix("", nrow = nrow(kinship_btp), ncol = ncol(kinship_btp))
rownames(celltext) <- rownames(kinship_btp)
colnames(celltext) <- colnames(kinship_btp)

celltext[paste0(male), paste0(female)] <- "&#9733;"

btp.plot <- heatmaply(
        kinship_btp,
        dendrogram            = "none", 
        xlab                  = "Females", 
        ylab                  = "Males",
        main                  = "Kinship Values for Potential Mate Pairs",
        scale                 = "none",
        colors                = paletteer_d("wesanderson::Zissou1Continuous"),
        margins               = c(60, 100, 40, 20),
        grid_color            = "white",
        grid_width            = 0.00001,
        label_format_fun      = function(value) round(value, digits = 3),
        cellnote              = celltext,
        cellnote_textposition = "middle center",
        cellnote_size         = 16,
        titleX                = TRUE,
        hide_colorbar         = FALSE,
        key.title             = "Kinship Vals",
        branches_lwd          = 0.1,
        fontsize_row          = 10, 
        fontsize_col          = 10,
        heatmap_layers        = theme(axis.line = element_blank())
        )

btp.plot
```



## IBD and Kappa Coefficients

Now we are going to print something called a Kappa Triangle. This is a way to visualize the most probable identity by descent (IBD) for this pair in a 2-dimensional space.

```{r}
kappa.pair <- coeff_living %>%
  filter(xor((id1 == male   & id2 == female), (id1 == female & id2 == male))) %>%
  select(id1, id2, kappa0, kappa1, kappa2)

kappa.tri <- showInTriangle(kappa.pair, plotType = "plotly")

print(kappa.pair)
kappa.tri
```

The red X mark represents the relationship between your hypothetical pair, while the letters in the graphic above are reference points for where different relationships would fall in that space. The summary table below contains a key for those abbreviations (and their associated values plotted here).

```{r}
print(example)
```

## Pedigree Plot

Now let's print that original pedigree once again with each of your individuals clearly marked.

```{r}
pair.ped  <- relabel(ped.living[[1]], 
                     new        = c("MATE_F", "MATE_M"),
                     old        = c(female  , male),
                     reorder    = TRUE,
                     returnLabs = TRUE)

 fills     <- list(
  "#A5AA99FF" = c(deceased),
  "#DAA51BFF" = c(founders),
  "#CC61B0FF" = c(female, "MATE_F"),
  "#2F8AC4FF" = c(male  , "MATE_M"),
  "#99C945FF" = c(undet)
)

plot(pair.ped,
     cex          = 0.5,
     deceased     = deceased,
     fill         = fills,
     lwd          = 0.8,
     font.lab     = 2,
     labs         = c("MATE_F", "MATE_M"),
     col.lab      = "black",
     starred      = c("MATE_F", "MATE_M"),
     hatched      = c("MATE_F", "MATE_M"),
     hatchDensity = 45,
     col          = "black",
     pty          = "m")
```


# Make your Plan and Report

Return to the report you began writing for the overall population statistics. Use all the information provided to match at least one breeding pair for your new exhibit and add a paragraph in which you explain this choice and plan.
