---
title: "Lab Assignment: Performing a PVA and Designing a BTP"
author: "Zoo Biology with Dr. Rich"
date: "Feb 11, 2025"
output:
  html_document:
    theme:
      bootswatch: litera
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: "hide"
    fig_caption: true
params:
  semester: "spring25"
  course: "zoobio"
                     
---

```{r, include = FALSE}
here::i_am("assignments/zoobio/labs/LabW4_BTP.Rmd")
global              <- config::get(config = "default")
templates           <- config::get(config = "templates")
term                <- config::get(config = params$semester)
course              <- config::get(config = params$course)
github              <- config::get(config = "github")
```


```{r, include = FALSE}
source(here::here(global$setup))
```

```{r, include = FALSE}
source(here("assignments/zoobio/labs/PVA/ui.R"))
source(here("assignments/zoobio/labs/PVA/server.R"))

# Run the application
shinyApp(ui = ui, server = server)
```

# Introduction

>Population Viability Analysis (PVA) is a key tool in zoo biology and conservation planning. It helps predict the future of a population by considering demographic trends, environmental conditions, and genetic factors. Today you will use real AZA studbook data for one species from your fictional zoo to conduct a PVA (Population Viability Analysis) and then design a BTP (Breeding and Transfer Plan) that matches SSP breeding pair(s) for your new facility to house.

In this assignment, you will:

- Convert a real **Species Survival Plan (SSP) studbook** (PDF) into an **Excel spreadsheet**.
- Upload your **studbook data** into a **Shiny app** to analyze **population viability**.
- Interpret **kinship matrices** and **genetic diversity metrics**.
- Select **individuals for transfer** to your group's new facility.
- Generate **breeding recommendations** based on kinship and genetic diversity.

---

# Assignment Instructions

## Step 1: Prepare your Studbook Spreadsheet and Select Candidate Individuals

```{r}
download_file(course$downloads$sample_studbook,
              "sample_studbook.xlsx",
              button_label   = "Download Studbook Template",
              button_type    = "info",
              self_contained = TRUE)
```


1. Download a **PDF studbook** assigned to your group.
2. Extract key data fields into an **Excel file** using the provided template (see above):
   - **Studbook ID**
   - **Current Location**
   - **House Name**
   - **Current Local ID**
   - **Sex Type** (M/F)
   - **Birth Date** (YYYY-MM-DD)
   - **Sire** (Father ID, if known)
   - **Dam** (Mother ID, if known)
3. Save your file as **Group_Studbook.xlsx**.
4. Look through your spreadsheet and deside on at least two individuals to propose as a candidate breeding pair for your facility.

---

## Step 2: Create a Pedigree for your Candidate Pair

1.	Open the QuickPed Shiny app: QuickPed.
2.	Manually enter pedigree data (IDs, sires, dams, sex).
3.	Download the pedigree file as a CSV.


## Step 3: Upload the Pedigree to the Shiny App
	
1.	Open the custom Shiny app (link provided by the instructor).
2.	Click Upload CSV and select your formatted pedigree file.
3.	Click Process Pedigree Data to:
  - Create the pedigree structure using pedtools.
  - Calculate kinship matrices.
  - Prepare for breeding recommendations.

## Step 2: Upload and Process Data in the Shiny App
1. Open the **PVA Shiny App** (click button below).
2. Click **Upload Studbook** and select your **Group_Studbook.xlsx** file.
3. Click **Process Studbook Data** to extract population parameters.
4. Review:
   - **Population size**
   - **Birth and death rates**
   - **Kinship matrix and genetic diversity**

---

## **Step 3: Run a Population Viability Analysis (PVA)**
1. Adjust **birth and death rates** based on your data.
2. Set a **carrying capacity** for the zoo facility.
3. Click **Run Simulation** to visualize **population trends**.
4. Take a screenshot of your **PVA graph** for your report.

---

## **Step 4: Select Individuals for Transfer**
1. Click **Show Kinship Matrix** to review genetic relationships.
2. Select **at least 3 individuals** with low mean kinship values.
3. Click **Confirm Transfer** for each selected individual.
4. Justify your selection in your report.

---

## **Step 5: Generate Breeding Recommendations**
1. Click **Generate Breeding Pairs**.
2. Download the **breeding recommendations report** (CSV/Excel).
3. Discuss:
   - Why were certain pairs selected?
   - How do these pairings help maintain genetic diversity?

---

# **Report Submission**
Your report should include:
- **Introduction** (What is PVA? Why is it important in Zoo Management?)
- **Methods** (How did you prepare and analyze the data?)
- **Results**
  - PVA simulation graph and interpretation
  - Kinship matrix and selected individuals
  - Breeding recommendations
- **Discussion**
  - What challenges did you encounter?
  - How could zoos improve population management?
    - Describe at least one metric or technique you would add or change to current population management protocols to improve the longterm viability and wellbeing of zoo populations.

**Submission Deadline:** `r Sys.Date() + 7`  
**Format:** PDF (submit one report per group on Canvas)  

---

# **Grading Criteria**
| Section                  | Points |
|--------------------------|--------|
| Introduction             | 2      |
| Data Preparation         | 3      |
| PVA Analysis & Graph     | 5      |
| Kinship & Transfer Plan  | 6      |
| Breeding Recommendations | 6      |
| Discussion & Insights    | 8      |
| **Total**                | **30**|

Good luck, and have fun analyzing zoo population sustainability! 🦁🐧🐍

