---
title: "Population Data Management"
author: "Alicia Rich"
output:
  html_document:
    theme:
      bootswatch: litera
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: "show"
    fig_caption: true
    df_print: paged
params:
  course: "zoobio"
  semester: "spring25"
                     
---

```{r, include = FALSE}
global             <- config::get(config = "default")
here::i_am("assignments/zoobio/labs/PopulationData.Rmd")
source(here::here(global$setup))

theme_set(theme_classic())
thematic_rmd()
thematic_on(accent = "#8785B2FF", fg = "black")

```


```{r}
psl.historic <- read.csv(here(
  paste0(course$assignments, "/testing/StudbookPSL2021Historic.csv")
  ), header = T) %>%
  mutate(across(where(is.character), ~ na_if(., "")),
         across(where(is.character), ~ str_trim(.)),
         Sex = if_else(Sex == "Female" | Sex == "Male", Sex, "Undetermined"),
         Location = str_remove_all(Location, "[^\\w]"))  %>%
       fill(ID,
            Current_Location,
            Current_Local_ID,
            Sex             ,
            Birth_Date      ,
            Sire            ,
            Dam             ,
            Birth_Type)  %>%
     select(ID,
            Current_Location,
            Sex             ,
            Sire            ,
            Dam             ,
            Birth_Type      ,
            Event_Type      ,
            Date            ,
            Event_Location = Location        ) %>%
  mutate(Event_Date   = dmy(str_extract(Date      , "\\d+/\\w+/\\d+")),
         Sire = as.integer(Sire),
         Dam  = as.integer(Dam),
         ID   = as.integer(ID),
         .keep = "unused")


psl.2021 <- read.csv(here(
  paste0(course$assignments, "/testing/StudbookPSL2021Living.csv")
  ), header = T) %>%
  mutate(across(where(is.character), ~ na_if(., "")),
         across(where(is.character), ~ str_trim(.)),
         Sex = if_else(Sex == "Female" | Sex == "Male", Sex, "Undetermined"),
         Location = str_remove_all(Location, "[^\\w]"))  %>%
       fill(ID,
            Current_Location,
            ID_Local,
            Sex             ,
            Birth_Date      ,
            Status          ,
            Sire            ,
            Dam             ,
            Birth_Type      ,
            Event_Type      ,
            Event_Date      ,
            Location) %>%
     select(ID,
            Current_Location,
            Sex             ,
            Status          ,
            Sire            ,
            Dam             ,
            Birth_Type      ,
            Event_Type      ,
            Event_Date      ,
            Event_Location = Location)  %>%
  mutate(Event_Date   = dmy(str_extract(Event_Date, "\\d+/\\w+/\\d+")),
         Sire = as.integer(Sire),
         Dam  = as.integer(Dam),
         ID   = as.integer(ID),
         .keep = "unused")

psl.2024 <- read.csv(here(
  paste0(course$assignments, "/testing/StudbookPSL2024Living.csv")
  ), header = T) %>%
  mutate(across(where(is.character), ~ na_if(., "")),
         Sex = if_else(Sex == "Female" | Sex == "Male", Sex, "Undetermined"))  %>%
       fill(ID,
            Current_Location,
            ID_Local,
            Sex             ,
            Status          ,
            Sire            ,
            Dam             ,
            Birth_Type      ,
            Event_Type      ,
            Event_Date      ,
            Location) %>%
  select(ID,
         Current_Location,
         Sex             ,
         Status          ,
         Sire            ,
         Dam             ,
         Birth_Type      ,
         Event_Type      ,
         Event_Date      ,
         Event_Location = Location
  ) %>%
  mutate(Event_Date   = dmy(str_extract(Event_Date, "\\d+/\\w+/\\d+")),
         Sire = as.integer(Sire),
         Dam  = as.integer(Dam),
         ID   = as.integer(ID),
         .keep = "unused")


```

```{r}
psl <- psl.2024 %>% 
  left_join(psl.2021) %>%
  left_join(psl.historic)
```

```{r}
psl.living <- psl.2024 %>% 
  left_join(psl.2021) %>%
  group_by(ID) %>% 
  mutate(Birth_Date     = if_else(str_detect(Event_Type, "Birth"), Event_Date, NA),
         Birth_Location = if_else(str_detect(Event_Type, "Birth"), Event_Location, NA)) %>%
  fill(Birth_Date, Birth_Location) %>%
  arrange(ID, Event_Date) %>%
  mutate(Event_Order = row_number()) %>%
  slice(max(Event_Order)) %>%
  ungroup() %>%
  select(ID,
         Current_Location,
         Sex,
         Sire,
         Dam,
         Birth_Type,
         Birth_Date,
         Birth_Location)
```



```{r}
psl.studbook <- psl.historic %>%
  mutate(Death_Date     = if_else(str_detect(Event_Type, "Death"), Event_Date, NA),
         Death_Location = if_else(str_detect(Event_Type, "Death"), Event_Location, NA),
         Birth_Location = if_else(str_detect(Event_Type, "Birth"), Event_Location, NA),
         Birth_Date     = if_else(str_detect(Event_Type, "Birth"), Event_Date, NA),
         Capture_Date   = if_else(str_detect(Event_Type, "Capture"), Event_Date, NA)) %>%
  fill(Death_Date,    
       Death_Location,
       Birth_Location,
       Birth_Date    ,
       Capture_Date  ) %>%
  group_by(ID)  %>%
  arrange(ID, Event_Date) %>%
  mutate(Event_Order = row_number()) %>%
  slice(max(Event_Order)) %>%
  ungroup() %>%
  select(ID,
         Current_Location,
         Sex,
         Sire,
         Dam,
         Birth_Type,
         Birth_Date,
         Birth_Location,
         Death_Date,    
         Death_Location,
       Capture_Date  ) %>%
  full_join(psl.living) %>%
  arrange(ID) %>%
  mutate(Sex = case_when(
    Sex == "Male"   ~ 1,
    Sex == "Female" ~ 2,
    .default = 3
  )) %>% distinct()
  
```



Introduction

In this tutorial, we will use a small example of studbook-style data to learn how to:
	1.	Import & clean demographic and genealogical data.
	2.	Create a pedigree object in R.
	3.	Visualize a pedigree.
	4.	Calculate basic population genetics metrics such as inbreeding coefficients and mean kinship.

Although real studbooks may contain thousands of rows with complex life-history, transfer, and breeding records, the methods demonstrated here scale to larger datasets.

Import and Inspect Data

Clean and Filter Data

Studbooks typically contain multiple rows per individual (transfer events, births, deaths, etc.). For pedigree creation, we usually just need:
	•	A single row per individual (often the “Birth/Hatch” row).
	•	The ID, sire, dam, and sex.
	•	(Optionally) a living/deceased flag if focusing on current living population metrics.

Below, we’ll select only Birth/Hatch rows and ensure we have one row per unique ID.

```{r}
df_births <- df_raw %>%
  # Filter to rows describing the actual birth/hatch event
  filter(Event_Type == "Birth/Hatch") %>%
  # If an individual has multiple "Birth/Hatch" lines for some reason, keep one
  distinct(ID, .keep_all = TRUE)

df_births
```

Next, let’s ensure the sex variable is in a format acceptable for pedtools. According to pedtools documentation, sex can be specified as:
	•	"male", "female", or "unknown"
	•	or numeric codes 1 (male), 2 (female), 3 (unknown).

We’ll do a simple recoding:

```{r}
df_births <- df_births %>%
  mutate(
    Sex = case_when(
      Sex == "Male" ~ "male",
      Sex == "Female" ~ "female",
      TRUE ~ "unknown"
    )
  )
```

Create a Pedigree Object

With pedtools, we can create a pedigree object by specifying:
	•	id:  vector of individual IDs
	•	fid: father IDs (sire)
	•	mid: mother IDs (dam)
	•	sex: each individual’s sex
	
```{r}
ped <- pedigree(
  id     = psl.studbook$ID,
  dadid    = psl.studbook$Sire,
  momid    = psl.studbook$Dam,
  sex    = psl.studbook$Sex
)
```


```{r}
# Check the pedigree
ped
```

Note: If you have any unknown sires/dams coded as NA or 0, pedtools will treat these as founders.

Visualize the Pedigree

To visualize the pedigree, you can use the built-in plot() method from pedtools. For larger pedigrees, you might want more advanced layout options, but for a small dataset, a simple plot can help illustrate relationships.

```{r}
plot(ped, id = ped$id)  # label individuals by their ID
```

Calculate Basic Genetic Metrics

Inbreeding Coefficients

The inbreeding coefficient (F) of an individual measures how likely it is that the two alleles at a locus in that individual are identical by descent. We can use inbreeding() from pedtools.

```{r}
# Calculate inbreeding for all individuals
inb <- inbreeding(ped)
inb
```

Pairwise Kinship Matrix

The kinship coefficient between two individuals is the probability that a random allele from each individual is identical by descent. We can calculate the full pairwise kinship matrix with:

```{r}
K <- kinship(ped)
K
```

	•	K[i, i] is half of the inbreeding coefficient for individual i.
	•	K[i, j] for i != j indicates the degree of relatedness.

Mean Kinship (MK)

Mean kinship for an individual is the average kinship that individual has with all living members of the population (including itself). A simple way to calculate mean kinship for each individual is:

```{r}
# We'll define "living" as individuals with Living == "Y" in df_births
living_ids <- df_births %>%
  filter(Living == "Y") %>%
  pull(ID)

# Subset the kinship matrix to only living individuals
K_living <- K[living_ids, living_ids, drop = FALSE]

# For each living individual, compute mean kinship across all living individuals
mk_values <- rowMeans(K_living)

# Turn it into a dataframe for clarity
mk_df <- data.frame(
  ID = living_ids,
  MeanKinship = mk_values
)

mk_df
```

Gene Diversity / Expected Heterozygosity

If you assume founders are unrelated and non-inbred, you can approximate gene diversity or expected heterozygosity as:

[
GD = 1 - \bar{F}
]

where (\bar{F}) is the average inbreeding coefficient of the living population. A quick example:

```{r}
# Average inbreeding among the living population:
avg_inb_living <- mean(inb[living_ids])

# Gene diversity:
GD <- 1 - avg_inb_living

GD
```

(Real-world SSP calculations often incorporate more complex assumptions about founder relationships, incomplete pedigrees, etc., so this is a simplified demonstration.)

Interpreting Results

	•	Pedigree Plot
The plotted pedigree provides a quick visual check for errors, unusual loops, missing parent data, etc.
	•	Inbreeding
A result of F = 0 indicates no inbreeding known in the pedigree, while higher values suggest some level of relatedness between the parents.
	•	Mean Kinship
Individuals with low mean kinship are considered genetically valuable for breeding (less related to the rest of the population).
	•	Gene Diversity
A summary statistic reflecting the population’s overall diversity as inferred from the pedigree.

Extensions

	•	Handling Unknown Founders: If your data has 0 or NA for unknown parents, those individuals become “founders” in the pedigree. Real founder relationships might not be truly random, so advanced models can incorporate partial knowledge or assumptions.
	•	Breeding Recommendations: In a real SSP, these metrics feed into management software (e.g., PMx or ZIMS). While R packages can replicate some calculations, official software remains the standard for final breeding recommendations.
	•	Larger Pedigrees: For a few thousand individuals, pedtools is usually quite fast. For very large pedigrees, you may need more memory-efficient methods or specialized tools.
	
	
Conclusion
	You now have a basic framework in R and R Markdown for:
	1.	Importing studbook data
	2.	Converting it to a pedigree object
	3.	Plotting family relationships
	4.	Calculating common metrics (inbreeding, kinship, mean kinship, and approximate gene diversity)