---
title: "Population Data Management"
author: "Alicia Rich"
output:
  html_document:
    theme:
      bootswatch: litera
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: "show"
    fig_caption: true
    df_print: paged
params:
  course: "zoobio"
  semester: "spring25"
                     
---

```{r, include = FALSE}
global             <- config::get(config = "default")
here::i_am("assignments/zoobio/labs/PopulationData.Rmd")
source(here::here(global$setup))

theme_set(theme_classic())
thematic_rmd()
thematic_on(accent = "#8785B2FF", fg = "black")

opts_chunk$set(message = FALSE,
               warning = FALSE,
               echo    = FALSE,
               include = TRUE,
               eval    = TRUE,
               comment = "")

source(here(paste0(course$labs, "ImportStudbooks.R")))
source(here(paste0(course$labs, "PedigreeFunctions.R")))
```

```{r, warning = FALSE}
studbook <- read_studbook_historic(here(paste0(course$assignments, "/testing/StudbookPSL2021Historic.csv"))) %>%
  full_join(read_studbook_current(here(paste0(course$assignments , "/testing/StudbookPSL2024Living.csv")))) %>%
  full_join(read_studbook_current(here(paste0(course$assignments , "/testing/StudbookPSL2021Living.csv"))))   %>%
  format_studbook(sex_unknown = 0) %>%
  sub_missing_parents()  %>%
  mutate(across(where(is.numeric), ~ replace_na(., 0)))
```


```{r}
living      <- studbook %>% filter(status == "alive")    %>% list_ids()
deceased    <- studbook %>% filter(status != "alive")    %>% list_ids()
sex.unknown <- studbook %>% filter(Sex == 0)             %>% list_ids()

studbook_colored <- studbook %>%
  mutate(fill = case_when(
    status == "deceased" | status == "Undetermined"  ~ "#A5AA99FF",
    status == "alive" & Sex == 0                     ~ "#99C945FF",
    status == "alive" & Sex == 1                     ~ "#2F8AC4FF",
    status == "alive" & Sex == 2                     ~ "#CC3A8EFF"
  ))
```


```{r}
fill_key <- select(studbook_colored, name = ID, value = fill) %>% deframe() %>% as.list()
```


```{r}
pedig <- ped(
  id     = studbook$ID,
  fid    = studbook$Sire,
  mid    = studbook$Dam,
  sex    = studbook$Sex
) %>% parentsBeforeChildren() %>% foundersFirst()
```

```{r}
singletons <- list_singletons(pedig)
families   <- related_ids(pedig)
```

```{r}
studbook.annotate <- studbook %>% mutate(
  lineage = case_when(ID %in% singletons ~ "singleton",
                      ID %in% families   ~ "family")
)
```

```{r}
pedigree_set <- keep(pedig, \(x) pedsize(x) > 1)  %>% imap(., \(x, idx) as.list(idx)) %>% list_flatten(name_spec = "")
```

```{r}
plot1 <- draw_pedigree(ped1)
```


```{r}
ped1 <- pedig[["_comp1"]]
ped2 <- pedig[["_comp10"]]
ped3 <- pedig[["_comp13"]]
ped4 <- pedig[["_comp25"]]

plot1 <- draw_pedigree(ped1)
plot2 <- draw_pedigree(ped2)
plot3 <- draw_pedigree(ped3)
plot4 <- draw_pedigree(ped4)

plot <- plotPedList(
            list(plot1,
                 plot2,
                 plot3,
                 plot4),
            framed = FALSE,
            
            )
```

```{r}
ped.living <- subset(pedig, living)
```



```{r}
generations(psl.ped)
hasUnbrokenLoops(psl.ped)
hasInbredFounders(psl.ped)
hasSelfing(psl.ped)
```

```{r}
inbred  <- enframe(inbreeding(psl.ped))
```



```{r}
kappaIBD <- kappaIBD(psl.ped)

```
```{r}
showInTriangle(kappaIBD)
```


```{r}
writePed(psl.ped, paste0(course$assignments, "/testing/PSLpedigree"))
```



Calculate Basic Genetic Metrics

Inbreeding Coefficients

The inbreeding coefficient (F) of an individual measures how likely it is that the two alleles at a locus in that individual are identical by descent. We can use inbreeding() from pedtools.

```{r}
# Calculate inbreeding for all individuals
inb <- inbreeding(ped)
inb
```

Pairwise Kinship Matrix

The kinship coefficient between two individuals is the probability that a random allele from each individual is identical by descent. We can calculate the full pairwise kinship matrix with:

```{r}
K <- kinship(ped)
K
```

	•	K[i, i] is half of the inbreeding coefficient for individual i.
	•	K[i, j] for i != j indicates the degree of relatedness.

Mean Kinship (MK)

Mean kinship for an individual is the average kinship that individual has with all living members of the population (including itself). A simple way to calculate mean kinship for each individual is:

```{r}
# We'll define "living" as individuals with Living == "Y" in df_births
living_ids <- df_births %>%
  filter(Living == "Y") %>%
  pull(ID)

# Subset the kinship matrix to only living individuals
K_living <- K[living_ids, living_ids, drop = FALSE]

# For each living individual, compute mean kinship across all living individuals
mk_values <- rowMeans(K_living)

# Turn it into a dataframe for clarity
mk_df <- data.frame(
  ID = living_ids,
  MeanKinship = mk_values
)

mk_df
```

Gene Diversity / Expected Heterozygosity

If you assume founders are unrelated and non-inbred, you can approximate gene diversity or expected heterozygosity as:

[
GD = 1 - \bar{F}
]

where (\bar{F}) is the average inbreeding coefficient of the living population. A quick example:

```{r}
# Average inbreeding among the living population:
avg_inb_living <- mean(inb[living_ids])

# Gene diversity:
GD <- 1 - avg_inb_living

GD
```

(Real-world SSP calculations often incorporate more complex assumptions about founder relationships, incomplete pedigrees, etc., so this is a simplified demonstration.)

Interpreting Results

	•	Pedigree Plot
The plotted pedigree provides a quick visual check for errors, unusual loops, missing parent data, etc.
	•	Inbreeding
A result of F = 0 indicates no inbreeding known in the pedigree, while higher values suggest some level of relatedness between the parents.
	•	Mean Kinship
Individuals with low mean kinship are considered genetically valuable for breeding (less related to the rest of the population).
	•	Gene Diversity
A summary statistic reflecting the population’s overall diversity as inferred from the pedigree.

Extensions

	•	Handling Unknown Founders: If your data has 0 or NA for unknown parents, those individuals become “founders” in the pedigree. Real founder relationships might not be truly random, so advanced models can incorporate partial knowledge or assumptions.
	•	Breeding Recommendations: In a real SSP, these metrics feed into management software (e.g., PMx or ZIMS). While R packages can replicate some calculations, official software remains the standard for final breeding recommendations.
	•	Larger Pedigrees: For a few thousand individuals, pedtools is usually quite fast. For very large pedigrees, you may need more memory-efficient methods or specialized tools.
	
	
Conclusion
	You now have a basic framework in R and R Markdown for:
	1.	Importing studbook data
	2.	Converting it to a pedigree object
	3.	Plotting family relationships
	4.	Calculating common metrics (inbreeding, kinship, mean kinship, and approximate gene diversity)