---
title: "Population Data Management Lab"
author: "Alicia Rich"
output:
  html_document:
    theme:
      bootswatch: litera
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: "show"
    fig_caption: true
    df_print: paged
params:
  course: "zoobio"
  semester: "spring25"
                     
---

```{r, include = FALSE}
global             <- config::get(config = "default")
here::i_am("assignments/zoobio/labs/PopulationData.Rmd")
source(here::here(global$setup))

theme_set(theme_classic())
thematic_rmd()
thematic_on(accent = "#8785B2FF", fg = "black")

opts_chunk$set(message = FALSE,
               warning = FALSE,
               echo    = FALSE,
               include = TRUE,
               eval    = TRUE,
               comment = "")

source(here(paste0(course$labs, "ImportStudbooks.R")))
source(here(paste0(course$labs, "PedigreeFunctions.R")))
```


# Format Studbook Data

First, we need to read in our studbook data and clean it up to create a *tidy* format of one master table.

1.  Open the file `Studbook_Living.csv` and make sure it has the following columns with the exact names listed.
```
ID  Sex    Sire    Dam   Location    Event_Date    Event_Type
```
  - It is fine if you have extra columns, and all the columns may be in any order, but your spreadsheet must include at least these columns with the exact name syntax.
  - Rename any columns as necessary, re-save as a `.csv`, and proceed.

2.  Repeat the same for the file `Studbook_Historic.csv` using the following list of columns.
```
ID  Sex    Sire    Dam   Current_Location   Date    Location    Event_Type
```

```{r, warning = FALSE}
studbook <- read_studbook_historic(here(paste0(course$assignments, "/testing/StudbookPSL2021Historic.csv"))) %>%
  full_join(read_studbook_current(here(paste0(course$assignments , "/testing/StudbookPSL2024Living.csv")))) %>%
  full_join(read_studbook_current(here(paste0(course$assignments , "/testing/StudbookPSL2021Living.csv"))))   %>%
  format_studbook(sex_unknown = 0) %>%
  sub_missing_parents()  %>%
  mutate(across(where(is.numeric), ~ replace_na(., 0)))
```

We will also perform a series of filtering steps to extract named list of IDs like this:

```{r, warning = FALSE}
living    <- filter(studbook, status == "alive") %>% 
  pull(ID)

```

Let's take a glimpse of our formatted studbook now.

```{r}
studbook.tbl.1 <- studbook %>%
  mutate(Birth_Date = if_else(is.na(Birth_Date), Capture_Date, Birth_Date),
         Sex        = case_when(Sex == 1 ~ "Male",
                                Sex == 2 ~ "Female",
                                Sex == 0 ~ "Unknown")) %>%
  arrange(ID) %>%
  select(
    ID,
    status,       
    Sire       ,     
    Dam   ,
    Current_Location,
    Sex,
    Birth_Type,
    Birth_Date,
    Birth_Location,
    Death_Date,
    Death_Location,
    Age
  ) %>%
  reactable(
    theme               = espn(),
    sortable            = TRUE,
    resizable           = TRUE,
    filterable          = TRUE,
    searchable          = TRUE,
    defaultPageSize     = 20,
    showPageSizeOptions = TRUE,
    highlight           = TRUE,
  columns = list(
       ID               = colDef(name = "Studbook ID"),
       status           = colDef(name = "Living/Deceased"),       
       Sire             = colDef(name = "Father") ,     
       Dam              = colDef(name = "Mother") ,
       Current_Location = colDef(name = "Current/Last Location"),
       Sex              = colDef(name = "Sex"),
       Birth_Type       = colDef(name = "Type of Birth"),
       Birth_Date       = colDef(name = "Date of Birth or Capture", 
                                 format = colFormat(date = TRUE)),
       Birth_Location   = colDef(name = "Location of Birth or Capture"),
       Death_Date       = colDef(name = "Date of Death", format = colFormat(date = TRUE)),
       Death_Location   = colDef(name = "Location of Death"),
       Age              = colDef(name = "Age Now or at Death")
   )
)

studbook.tbl.1
```


# Construct Pedigree

Now we will use a package called `pedtools` to create our pedigree objects.

```{r}
pedig <- ped(
  id     = studbook$ID,
  fid    = studbook$Sire,
  mid    = studbook$Dam,
  sex    = studbook$Sex
)

founders    <- founders(pedig)
nonfounders <- nonfounders(pedig)
```

## Add to Summary Table

Let's look at our studbook with pedigree information added to it.

```{r}
studbook.tbl.2 <- studbook %>%
  mutate(Birth_Date = if_else(is.na(Birth_Date), Capture_Date, Birth_Date),
         Sex        = case_when(Sex == 1 ~ "Male",
                                Sex == 2 ~ "Female",
                                Sex == 0 ~ "Unknown"),
         Founder    = if_else(ID %in% as.integer(founders), "Yes", "No")) %>%
  arrange(ID) %>%
  select(
    ID,
    Founder,
    status,       
    Sire       ,     
    Dam   ,
    Current_Location,
    Sex,
    Birth_Type,
    Birth_Date,
    Birth_Location,
    Death_Date,
    Death_Location,
    Age
  ) %>%
  reactable(
    theme               = espn(),
    sortable            = TRUE,
    resizable           = TRUE,
    filterable          = TRUE,
    searchable          = TRUE,
    defaultPageSize     = 20,
    showPageSizeOptions = TRUE,
    highlight           = TRUE,
  columns = list(
       ID               = colDef(name = "Studbook ID"),
       Founder          = colDef(name = "Founder?") ,
       status           = colDef(name = "Living/Deceased"),       
       Sire             = colDef(name = "Father") ,     
       Dam              = colDef(name = "Mother") ,
       Current_Location = colDef(name = "Current/Last Location"),
       Sex              = colDef(name = "Sex"),
       Birth_Type       = colDef(name = "Type of Birth"),
       Birth_Date       = colDef(name = "Date of Birth or Capture", 
                                 format = colFormat(date = TRUE)),
       Birth_Location   = colDef(name = "Location of Birth or Capture"),
       Death_Date       = colDef(name = "Date of Death", format = colFormat(date = TRUE)),
       Death_Location   = colDef(name = "Location of Death"),
       Age              = colDef(name = "Age Now or at Death")
   )
)

studbook.tbl.2
```


```{r}
deceased    <- filter(studbook, status != "alive") %>% 
  pull(ID) %>% 
  setdiff(founders)

female    <- filter(studbook, status == "alive") %>% 
  pull(ID) %>% 
  intersect(nonfounders(pedig)) %>% 
  intersect(females(pedig))

male      <- filter(studbook, status == "alive") %>% 
  pull(ID) %>% 
  intersect(nonfounders(pedig)) %>% 
  intersect(males(pedig))

undet <- nonfounders(pedig) %>% 
  setdiff(females(pedig)) %>% 
  setdiff(males(pedig)) %>% 
  setdiff(deceased)

fills         <- list(
  "#A5AA99FF" = c(deceased),
  "#DAA51BFF" = c(founders),
  "#CC61B0FF" = c(female),
  "#2F8AC4FF" = c(male),
  "#99C945FF" = c(undet)
)
```

Now, we want to visualize a pedigree for just the representation of our living population, so we will subset the pedigree list and plot below.

```{r}
pedig.max   <- get_biggest_ped(pedig)
pedig.minor <- get_minor_peds(pedig)

twins.include <- subset_twins(studbook, pedig.max)
twins.list    <- twins_vector(twins.include)

  plot(pedig.max,
       cex      = 0.5,
       labs     = NULL,
       deceased = deceased,
       fill     = fills,
       col      = "black",
       twins    = twins.include)
```

# Generate Detailed Pedigree Statistics

```{r}
list_living <- as.list(living)

founder_reps <- map(list_living, \(x) as.list(ancestors(pedig, x))) %>%
  list_c() %>%
  enframe(name = NULL, value = "ID") %>% distinct() %>%
  filter(ID %in% founders(pedig)) %>%
  mutate(ID = as.integer(ID)) %>%
  left_join(studbook, by = join_by(ID))

n_founder_reps <- length(unique(founder_reps$ID))

living       <- filter(studbook, status == "alive") %>% 
  pull(ID) %>% 
  as.character()

livingIDs    <- intersect(nonfounders(pedig[["_comp1"]]), living)

founderIDs   <- founders(pedig[["_comp1"]])
dp           <- descentPaths(pedig[["_comp1"]], ids = founderIDs, internal = FALSE)
```

## View Progress

Now we will look at our studbook again with more information added.

```{r}
studbook.data <- studbook %>%
  mutate(Birth_Date = if_else(is.na(Birth_Date), Capture_Date, Birth_Date),
         Sex        = case_when(Sex == 1 ~ "Male",
                                Sex == 2 ~ "Female",
                                Sex == 0 ~ "Unknown"),,
         Founder     = if_else(ID %in% founders, "Yes", "No")) %>%
  mutate(Founder_rep = case_when(ID %in% founder_reps$ID ~ "Yes", 
                                  Founder == "No" ~ NA,
                                 .default = "No"))

studbook.tbl.3 <- studbook.data %>%
  arrange(ID) %>%
  select(
    ID,
    Founder,
    Founder_rep,
    status,       
    Sire       ,     
    Dam   ,
    Current_Location,
    Sex,
    Birth_Type,
    Birth_Date,
    Birth_Location,
    Death_Date,
    Death_Location,
    Age
  ) %>%
  reactable(
    theme               = espn(),
    sortable            = TRUE,
    resizable           = TRUE,
    filterable          = TRUE,
    searchable          = TRUE,
    defaultPageSize     = 20,
    showPageSizeOptions = TRUE,
    highlight           = TRUE,
  columns = list(
       ID               = colDef(name = "Studbook ID"),
       Founder          = colDef(name = "Founder?") ,
       Founder_rep      = colDef(name = "Founder Still Represented?") ,
       status           = colDef(name = "Living/Deceased"),       
       Sire             = colDef(name = "Father") ,     
       Dam              = colDef(name = "Mother") ,
       Current_Location = colDef(name = "Current/Last Location"),
       Sex              = colDef(name = "Sex"),
       Birth_Type       = colDef(name = "Type of Birth"),
       Birth_Date       = colDef(name = "Date of Birth or Capture", 
                                 format = colFormat(date = TRUE)),
       Birth_Location   = colDef(name = "Location of Birth or Capture"),
       Death_Date       = colDef(name = "Date of Death", format = colFormat(date = TRUE)),
       Death_Location   = colDef(name = "Location of Death"),
       Age              = colDef(name = "Age Now or at Death")
   )
)

studbook.tbl.3
```


## Calculate Kinship and Other Genetic Stats

```{r}
studbook.kin <- studbook %>%
  mutate(status = if_else(status != "alive", 1, 0),
         Sex    = if_else(Sex == 0, 3, Sex))

all.twins <- subset_twins(studbook, pedig)

pedig.kin <- pedigree(
  id       = studbook.kin$ID,
  dadid    = studbook.kin$Sire,
  momid    = studbook.kin$Dam,
  sex      = studbook.kin$Sex,
  status   = studbook.kin$status,
  relation = as.matrix(all.twins)
)


kinship_all    <- kinship(pedig.kin)

living         <- filter(studbook, status == "alive") %>% 
  mutate(ID = as.character(ID)) %>% pull(ID)

living.kin     <- intersect(living, rownames(kinship_all))

kinship_living <- kinship_all[living.kin, living.kin]

founderContribution <- map_dbl(founderIDs, function(f) {
  paths_list <- dp[[f]]
  valid_paths <- keep(paths_list, ~ tail(.x, 1) %in% livingIDs)
  sum(map_dbl(valid_paths, ~ 0.5^(length(.x) - 1)))
})
names(founderContribution) <- founderIDs

p <- founderContribution / sum(founderContribution)
FGE <- 1 / sum(p^2)

GD <- 1 - sum(p^2)

MK <- mean(kinship_living[upper.tri(kinship_living)])

F_vec <- 2 * diag(kinship_living) - 1
F_mean <- mean(F_vec)

gen_numbers <- generations(pedig[["_comp1"]], what = "indiv")

living_gen  <- gen_numbers[names(gen_numbers) %in% livingIDs]

mean_gen    <- mean(living_gen, na.rm = TRUE)

kin_mat <- kinship_living

delta_F <- 1 - (1 - F_mean)^(1/mean_gen)

# Compute effective population size (Ne) from ΔF:
Ne <- 1 / (2 * delta_F)

# Census size is simply the number of living individuals:
N <- length(livingIDs)
Ne_over_N <- Ne / N


studbook.revised <- studbook %>%
  mutate(Birth_Date = if_else(is.na(Birth_Date), Capture_Date, Birth_Date)) %>%
  arrange(Birth_Date) %>%
  mutate(Birth_Date = if_else(is.na(Birth_Date), first(Birth_Date), Birth_Date)) %>%
  filter(!is.na(ID))

max_year <- year(Sys.Date())

alive_years <- pmap(
  list(birth = studbook.revised$Birth_Date, death = studbook.revised$Death_Date, status = studbook.revised$status),
  function(birth, death, status) {
    if (is.na(birth)) return(NULL)
    start_year <- year(birth)
    if (is.na(death) && status == "alive") {
      end_year <- max_year
    } else if (is.na(death)) {
      return(NULL)
    } else {
      end_year <- year(death)
    }
    seq(start_year, end_year)
  }
)

all_years <- unlist(alive_years)

annual_counts <- table(all_years) %>% 
  as.data.frame(stringsAsFactors = FALSE) %>% 
  rename(Year = all_years, Count = Freq) %>%
  mutate(Year = as.integer(Year)) %>%
  arrange(Year)


growth_rates <- annual_counts$Count[-1] / annual_counts$Count[-length(annual_counts$Count)]

last_5_growth <- tail(growth_rates, 5)

lambda  <- prod(last_5_growth)^(1/length(last_5_growth))

lambda2 <- exp(mean(log(last_5_growth)))
```

## Visualize Summary Table

Now you can view a summary of the statistics we just calculated and compare them to the summaries provided in the example BTP. Use this table to write your own version of a summary (1-2 paragraphs max) like the example in front of you.

```{r}
pva.summary <- tibble(N, 
                      n_founder_reps, 
                      FGE, 
                      delta_F, 
                      Ne, 
                      Ne_over_N, 
                      mean_gen, 
                      lambda2, 
                      F_mean, 
                      MK, 
                      GD) %>%
  mutate(across(where(is.numeric), ~round(., digits = 3))) %>%
  reactable(
    columns = list(
      N                = colDef(header = tippy("N"          , tooltip = "Demographic Population size: Number of currently living individuals in the population"                          )), 
      Ne               = colDef(header = tippy("Ne"         , tooltip = "Effective population size (estimated from the rate of inbreeding per generation)"                            ),
                                format = colFormat(digits = 0)), 
      Ne_over_N        = colDef(header = tippy("Ne/N"       , tooltip = "Effective population size relative to population size"),
                                format = colFormat(digits = 1)), 
      n_founder_reps   = colDef(header = tippy("Founders"   , tooltip = "Number of founders represented in the current population"                                 )), 
      FGE              = colDef(header = tippy("FGE"        , tooltip = "Founder genome equivalents: expressed in units of the number of wild-caught, unrelated individuals (“founders”) that would produce the same gene diversity as the current population"                           ),
                                format = colFormat(digits = 1)), 
      F_mean           = colDef(header = tippy("F"          , tooltip = "Mean inbreeding level"                                      )), 
      mean_gen         = colDef(header = tippy("Generations", tooltip = "Mean generation for living individuals"               ),
                                format = colFormat(digits = 1)), 
      delta_F          = colDef(header = tippy("ΔF"         , tooltip = "ΔF is the per–generation increase in inbreeding"                    )), 
      GD               = colDef(header = tippy("GD %"       , tooltip = "Gene diversity: current percentage of founding gene diversity retained"                                       ),
                                format = colFormat(percent = TRUE)), 
      MK               = colDef(header = tippy("MK"         , tooltip = "Population mean kinship"                              )), 
      lambda2          = colDef(header = tippy("lambda"     , tooltip = "Growth rate: Finite rate of population growth calculated as the geometric mean of changes in total population size over the past 5 years"                                          ))
    )
  )

pva.summary
```

# Kinship Plots

## All Living Individuals

The plot below gives us a basic visualization of the overall patterns in relatedness between all living individuals in the population.

```{r}
kin.plot <- heatmaply(
        kinship_living,
        dendrogram     = "none",
        xlab           = "", ylab = "",
        main           = "",
        scale          = "none",
        colors         = paletteer_d("beyonce::X51"),
        margins        = c(60,100,40,20),
        grid_color     = "white",
        grid_width     = 0.00001,
        titleX         = FALSE,
        hide_colorbar  = TRUE,
        branches_lwd   = 0.1,
        fontsize_row   = 5, fontsize_col = 5,
        labCol         = colnames(kinship_living),
        labRow         = rownames(kinship_living),
        heatmap_layers = theme(axis.line = element_blank())
        )
```

## Matched Pairs

Finally, we want to look closer at potential breeding pairs to do some match-making, so let's generate some more pedigree statistics per individual and then visualize a plot of breeding pairs.


```{r}
ancestors <- as.list(rownames(kinship_living)) %>%
  set_names(map(., \(x) x)) %>%
  map(., \(x) as.list(ancestors(pedig[["_comp1"]], x))) %>%
  enframe(name = "ID", value = "Ancestors")

children <- as.list(rownames(kinship_living)) %>%
  set_names(map(., \(x) x)) %>%
  map(., \(x) as.list(children(pedig[["_comp1"]], x))) %>%
  enframe(name = "ID", value = "Children")

descendants <- as.list(rownames(kinship_living)) %>%
  set_names(map(., \(x) x)) %>%
  map(., \(x) as.list(descendants(pedig[["_comp1"]], x))) %>%
  enframe(name = "ID", value = "Descendants")

siblings <- as.list(rownames(kinship_living)) %>%
  set_names(map(., \(x) x)) %>%
  map(., \(x) as.list(siblings(pedig[["_comp1"]], x))) %>%
  enframe(name = "ID", value = "Siblings")

family.basics <- ancestors %>% 
  left_join(descendants) %>% 
  left_join(children) %>% 
  left_join(siblings) %>%
  rowwise() %>%
  mutate(N_Ancestors   = length(Ancestors),
         N_Children = length(Children),
         N_Descendants = length(Descendants),
         N_Siblings    = length(Siblings))

living.m       <- filter(studbook, status == "alive" & Sex == 1) %>% 
  mutate(ID = as.character(ID)) %>% 
  pull(ID)
living.f       <- filter(studbook, status == "alive" & Sex == 2) %>% 
  mutate(ID = as.character(ID)) %>% 
  pull(ID)

annotate <- select(family.basics, c(ID, starts_with("N_"))) %>%
  left_join(enframe(F_vec, name = "ID", value = "F_vec")) %>%
  left_join(enframe(gen_numbers, name = "ID", value = "gen_numbers")) %>%
  mutate(ID = as.integer(ID)) %>%
  left_join((studbook.data)) %>%
  mutate(ID = as.character(ID),
    hoverText = str_glue(
    "{Current_Location}", "<br>", 
    "{Age}", " yrs (Born ", "{Birth_Date}", ")<br>",
    "Gen Number: ", "{gen_numbers}", 
    ", Mother: ", "{Dam}", ", Father: ", "{Sire}", "<br>",
    "N Ancestors: ", "{N_Ancestors}", 
    ", N Siblings: ", "{N_Siblings}",
    ", N Children: ", "{N_Children}", 
    ", N Descendants: ", "{N_Descendants}"
    )
  ) %>% arrange(ID)


living_males   <- intersect(living.m, rownames(kinship_living))
living_females <- intersect(living.f, rownames(kinship_living))

kinship_btp    <- kinship_living[living_males, living_females]

text_males   <- intersect(living.m, rownames(kinship_btp))
text_females <- intersect(living.f, colnames(kinship_btp))

annotate.m   <- filter(annotate, status == "alive" & Sex == "Male") %>% 
  select(ID, hoverText) %>%
  filter(ID %in% text_males) %>% 
  arrange(match(ID, text_males))

annotate.f   <- filter(annotate, status == "alive" & Sex == "Female") %>% 
  select(ID, hoverText) %>%
  filter(ID %in% text_females) %>% 
  arrange(match(ID, text_females))

hover_matrix <- outer(
  annotate.m$hoverText, 
  annotate.f$hoverText, 
  FUN = function(m_text, f_text) {
    I(paste0("<br>Male:<br>", m_text, "<br><br>",
           "Female:<br>", f_text))
  }
)

rownames(hover_matrix) <- annotate.m$ID
colnames(hover_matrix) <- annotate.f$ID

btp.plot <- heatmaply(
        kinship_btp,
        dendrogram       = "none", 
        xlab             = "Females", 
        ylab             = "Males",
        main             = "Kinship Values for Potential Mate Pairs",
        scale            = "none",
        colors           = paletteer_d("beyonce::X51"),
        margins          = c(60, 100, 40, 20),
        grid_color       = "white",
        grid_width       = 0.00001,
        label_format_fun = function(value) round(value, digits = 3),
        titleX           = TRUE,
        hide_colorbar    = FALSE,
        key.title        = "Kinship Vals",
        branches_lwd     = 0.1,
        fontsize_row     = 10, 
        fontsize_col     = 10,
        custom_hovertext = I(hover_matrix),
        heatmap_layers   = theme(axis.line = element_blank())
        )

btp.plot
```

# Make your Plan and Report

Return to the report you began writing for the overall population statistics. Use all the information provided to match at least one breeding pair for your new exhibit and add a paragraph in which you explain this choice and plan.
