---
title: "Population Data Management"
author: "Alicia Rich"
output:
  html_document:
    theme:
      bootswatch: litera
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: "show"
    fig_caption: true
    df_print: paged
params:
  course: "zoobio"
  semester: "spring25"
                     
---

```{r, include = FALSE}
global             <- config::get(config = "default")
here::i_am("assignments/zoobio/labs/PopulationData.Rmd")
source(here::here(global$setup))

theme_set(theme_classic())
thematic_rmd()
thematic_on(accent = "#8785B2FF", fg = "black")

opts_chunk$set(message = FALSE,
               warning = FALSE,
               echo    = FALSE,
               include = TRUE,
               eval    = TRUE,
               comment = "")

```

```{r}
read_studbook_current <- function(file) {
  read.csv(file, header = T) %>%
  mutate(across(where(is.character), ~ na_if(., "")),
         across(where(is.character), ~ str_trim(.)),
         across(where(is.character), ~ str_remove_all(., "\n")),
         Sex = if_else(Sex == "Female" | Sex == "Male", Sex, "Undetermined"),
         Event_Location = str_remove_all(Location, "[^\\w]"),
         Event_Date   = dmy(str_extract(Event_Date, "\\d+/\\w+/\\d+"))) %>%
    mutate(
         Sire = as.integer(Sire),
         Dam  = as.integer(Dam),
         ID   = as.integer(ID))  %>%
  mutate(Birth_Date     = if_else(str_detect(Event_Type, "Birth"), Event_Date, NA),
         Birth_Location = if_else(str_detect(Event_Type, "Birth"), Event_Location, NA),
         status         = "alive") %>%
    fill(ID) %>%
    group_by(ID) %>%
       fill(Current_Location,
            ID_Local,
            Sex             ,
            status          ,
            Sire            ,
            Dam             ,
            Birth_Type      ,
            Birth_Date      ,
            Event_Type      ,
            Event_Date      ,
            Event_Location  ,
            Birth_Location) %>%
    ungroup()  %>%
    select(
      ID           ,
      Current_Location,
      Sex       ,
      status     ,
      Birth_Date   ,
      Sire   ,
      Dam,
      Birth_Type     ,
      Event_Type     ,
      Event_Location ,
      Event_Date ,
      Birth_Location
    )
}

read_studbook_historic <- function(file) {
  read.csv(file, header = T) %>%
  mutate(across(where(is.character), ~ na_if(., "")),
         across(where(is.character), ~ str_trim(.)),
         across(where(is.character), ~ str_remove_all(., "\n")),
         Sex      = if_else(Sex == "Female" | Sex == "Male", Sex, "Undetermined"),
         Event_Location = str_remove_all(Location, "[^\\w]"), 
         Event_Date   = dmy(str_extract(Date, "\\d+/\\w+/\\d+")),
         status         = "deceased")  %>%
    mutate(
         Sire = as.integer(Sire),
         Dam  = as.integer(Dam),
         ID   = as.integer(ID))  %>%
  mutate(Death_Date     = if_else(str_detect(Event_Type, "Death"), Event_Date, NA),
         Death_Location = if_else(str_detect(Event_Type, "Death"), Event_Location, NA),
         Birth_Location = if_else(str_detect(Event_Type, "Birth"), Event_Location, NA),
         Birth_Date     = if_else(str_detect(Event_Type, "Birth"), Event_Date, NA),
         Capture_Date   = if_else(str_detect(Event_Type, "Capture"), Event_Date, NA)) %>%
    fill(ID) %>%
    group_by(ID) %>%
       fill(Current_Location,
            Sex             ,
            Sire            ,
            Dam             ,
            Birth_Type,
            Death_Date,    
            Death_Location,
            Birth_Location,
            Birth_Date    ,
            Capture_Date  )  %>%
    ungroup() %>%
    select(
      ID           ,
      Current_Location,
      Sex       ,
      status       ,
      Birth_Date   ,
      Sire   ,
      Dam,
      Birth_Type     ,
      Event_Type     ,
      Event_Location ,
      Event_Date ,
      Death_Date,
      Death_Location,
      Birth_Location,
      Capture_Date
    )
}

sub_missing_parents <- function(df) {
  df %>%
  bind_rows(tibble(ID               = c(setdiff(unique(df$Sire), df$ID)),
                          Current_Location = NA_character_,
                          Sex              = 1,
                          status           = "Undetermined",
                          Sire             = NA_integer_,
                          Dam              = NA_integer_,
                          Birth_Type       = "Undetermined",
                          Birth_Date       = as.Date(NA),
                          Birth_Location   = "Undetermined",
                          Death_Date       = as.Date(NA),
                          Death_Location   = "Undetermined",
                          Capture_Date     = as.Date(NA)
                                                  ), 
            tibble(ID               = c(setdiff(unique(df$Dam), df$ID)),
                          Current_Location = NA_character_,
                          Sex              = 2,
                          status           = "Undetermined",
                          Sire             = NA_integer_,
                          Dam              = NA_integer_,
                          Birth_Type       = "Undetermined",
                          Birth_Date       = as.Date(NA),
                          Birth_Location   = "Undetermined",
                          Death_Date       = as.Date(NA),
                          Death_Location   = "Undetermined",
                          Capture_Date     = as.Date(NA)
                                                  )
            )  %>%
  filter(!is.na(ID))
}

```


```{r}
format_studbook <- function(df, sex_unknown) {
  df   %>%
  filter(!is.na(Event_Date)) %>%
  arrange(ID, Event_Date) %>%
  group_by(ID) %>%
  mutate(Event_Order = row_number()) %>%
  slice(max(Event_Order)) %>%
  ungroup() %>%
  mutate(Sex = case_when(
    Sex == "Male"   ~ 1,
    Sex == "Female" ~ 2,
    .default = sex_unknown
  )) %>%
  select(ID,
         Current_Location,
         status,
         Sex,
         Sire,
         Dam,
         Birth_Type,
         Birth_Date,
         Birth_Location,
         Death_Date,    
         Death_Location,
         Capture_Date  ) %>%
  arrange(ID) %>%
  filter(!(str_detect(Birth_Type, "Captive") & (is.na(Sire) | is.na(Dam)))) %>%
  filter(!is.na(ID)) %>%
  mutate(Age = case_when(status == "alive"                         ~ year(today())    - year(Birth_Date), 
                         status == "deceased" & !is.na(Birth_Date) ~ year(Death_Date) - year(Birth_Date),
                         status == "deceased" & is.na(Birth_Date) & !is.na(Capture_Date) ~ year(Death_Date) - year(Capture_Date),
         .default = 0))
}
```

```{r, warning = FALSE}
studbook <- read_studbook_historic(here(paste0(course$assignments, "/testing/StudbookPSL2021Historic.csv"))) %>%
  full_join(read_studbook_current(here(paste0(course$assignments , "/testing/StudbookPSL2024Living.csv")))) %>%
  full_join(read_studbook_current(here(paste0(course$assignments , "/testing/StudbookPSL2021Living.csv"))))   %>%
  format_studbook(sex_unknown = 0) %>%
  sub_missing_parents()  %>%
  mutate(across(where(is.numeric), ~ replace_na(., 0)))
```


```{r}
living   <- studbook %>% filter(status == "alive")    %>% distinct(ID) %>% map(\(x) as.list(x)) %>% list_flatten(name_spec = "")

deceased <- studbook %>% filter(status == "deceased") %>% distinct(ID) %>% map(\(x) as.list(x)) %>% list_flatten(name_spec = "")
```


```{r}
pedig <- ped(
  id     = studbook$ID,
  fid    = studbook$Sire,
  mid    = studbook$Dam,
  sex    = studbook$Sex
)
```

```{r}
singletons <- keep(pedig, \(x) pedsize(x) <= 1) %>% map_depth(., 1, \(x) as.list(x[["ID"]])) %>% list_flatten(name_spec = "")

families   <- keep(pedig, \(x) pedsize(x) > 1)  %>% map_depth(., 1, \(x) as.list(x[["ID"]])) %>% list_flatten(name_spec = "")
```

```{r}
studbook.annotate <- studbook %>% mutate(
  lineage = case_when(ID %in% singletons ~ "singleton",
                      ID %in% families   ~ "family")
)
```




```{r}
psl.plot <- plot(psl.ped[["_comp3"]], labs = NULL)
```

```{r}
psl.plot <- plotPedList(psl.ped, labs = NULL)
```


```{r}
pedsize(psl.ped)
```

```{r}
generations(psl.ped)
hasUnbrokenLoops(psl.ped)
hasInbredFounders(psl.ped)
hasSelfing(psl.ped)
```

```{r}
inbred  <- enframe(inbreeding(psl.ped))
```

```{r}
psl <- ped(
  id     = psl.studbook$ID,
  fid    = psl.studbook$Sire,
  mid    = psl.studbook$Dam,
  sex    = psl.studbook$Sex
)
```


```{r}
kappaIBD <- kappaIBD(psl.ped)

```
```{r}
showInTriangle(kappaIBD)
```


```{r}
writePed(psl.ped, paste0(course$assignments, "/testing/PSLpedigree"))
```



Calculate Basic Genetic Metrics

Inbreeding Coefficients

The inbreeding coefficient (F) of an individual measures how likely it is that the two alleles at a locus in that individual are identical by descent. We can use inbreeding() from pedtools.

```{r}
# Calculate inbreeding for all individuals
inb <- inbreeding(ped)
inb
```

Pairwise Kinship Matrix

The kinship coefficient between two individuals is the probability that a random allele from each individual is identical by descent. We can calculate the full pairwise kinship matrix with:

```{r}
K <- kinship(ped)
K
```

	•	K[i, i] is half of the inbreeding coefficient for individual i.
	•	K[i, j] for i != j indicates the degree of relatedness.

Mean Kinship (MK)

Mean kinship for an individual is the average kinship that individual has with all living members of the population (including itself). A simple way to calculate mean kinship for each individual is:

```{r}
# We'll define "living" as individuals with Living == "Y" in df_births
living_ids <- df_births %>%
  filter(Living == "Y") %>%
  pull(ID)

# Subset the kinship matrix to only living individuals
K_living <- K[living_ids, living_ids, drop = FALSE]

# For each living individual, compute mean kinship across all living individuals
mk_values <- rowMeans(K_living)

# Turn it into a dataframe for clarity
mk_df <- data.frame(
  ID = living_ids,
  MeanKinship = mk_values
)

mk_df
```

Gene Diversity / Expected Heterozygosity

If you assume founders are unrelated and non-inbred, you can approximate gene diversity or expected heterozygosity as:

[
GD = 1 - \bar{F}
]

where (\bar{F}) is the average inbreeding coefficient of the living population. A quick example:

```{r}
# Average inbreeding among the living population:
avg_inb_living <- mean(inb[living_ids])

# Gene diversity:
GD <- 1 - avg_inb_living

GD
```

(Real-world SSP calculations often incorporate more complex assumptions about founder relationships, incomplete pedigrees, etc., so this is a simplified demonstration.)

Interpreting Results

	•	Pedigree Plot
The plotted pedigree provides a quick visual check for errors, unusual loops, missing parent data, etc.
	•	Inbreeding
A result of F = 0 indicates no inbreeding known in the pedigree, while higher values suggest some level of relatedness between the parents.
	•	Mean Kinship
Individuals with low mean kinship are considered genetically valuable for breeding (less related to the rest of the population).
	•	Gene Diversity
A summary statistic reflecting the population’s overall diversity as inferred from the pedigree.

Extensions

	•	Handling Unknown Founders: If your data has 0 or NA for unknown parents, those individuals become “founders” in the pedigree. Real founder relationships might not be truly random, so advanced models can incorporate partial knowledge or assumptions.
	•	Breeding Recommendations: In a real SSP, these metrics feed into management software (e.g., PMx or ZIMS). While R packages can replicate some calculations, official software remains the standard for final breeding recommendations.
	•	Larger Pedigrees: For a few thousand individuals, pedtools is usually quite fast. For very large pedigrees, you may need more memory-efficient methods or specialized tools.
	
	
Conclusion
	You now have a basic framework in R and R Markdown for:
	1.	Importing studbook data
	2.	Converting it to a pedigree object
	3.	Plotting family relationships
	4.	Calculating common metrics (inbreeding, kinship, mean kinship, and approximate gene diversity)